---
title: "Internet Connectivity and Usage Across the United States"
author: "Robert Jurgens, Logan Watkins, Thu Le"
output:
  html_document:
    css: styles.css
  html_notebook:
    css: styles.css
---
```{r}

#  _____         _                  _        _                                   _______  _            
# |  __ \       | |                | |      | |                                 |__   __|| |           
# | |__) | ___  | |__    ___  _ __ | |_     | |      ___    __ _   __ _  _ __      | |   | |__   _   _ 
# |  _  / / _ \ | '_ \  / _ \| '__|| __|    | |     / _ \  / _` | / _` || '_ \     | |   | '_ \ | | | |
# | | \ \| (_) || |_) ||  __/| |   | |_  _  | |____| (_) || (_| || (_| || | | | _  | |   | | | || |_| |
# |_|  \_\\___/ |_.__/  \___||_|    \__|( ) |______|\___/  \__, | \__,_||_| |_|( ) |_|   |_| |_| \__,_|
#                                       |/                  __/ |              |/                         
#                                                          |___/               
```


```{r setup, include=FALSE}
source("Required.R")
source("getToken.R")
library(knitr)
opts_chunk$set(out.width='1750px', dpi=200)
```

# I. Session Info
```{r}
sessionInfo()
```

# II. Data Sources

Note : set the working directory to the 00 Doc Folder.

<b>Original data sources:</b></br>
http://www.governing.com/gov-data/internet-usage-by-state.html</br>
https://www.ntia.doc.gov/data/digital-nation-data-explorer#sel=tvBoxUser&disp=map

To download the clean data (what was used in Tableau) download CleanedInternetUsageBySatet.csv from the <b>data.world</b> link.
https://data.world/lowatt/s-17-dv-project-6 and download ASM_2015_31AS101_with_ann.csv.

This data describes Internet Usage and Connectivity across the United States from July 2015 to October 2016. The columns that we used were a combination of the census population data,income data, internet usage and connectivity data.

# III. ETL Process

## 1. Bringing in the data
First, we brought the US Census data into our R environment.

```{r, warning=FALSE}
df1 <- data.world::query(connection = conn,
                         type = "sql",
                         dataset = "thule179/s-17-dv-final-project",
                         "select 
                         A.AreaName,
                         A.`B01001_001`, 
                         A.`B01001_002`,
                         A.`B01001_003`,
                         A.`B01001_004`,
                         A.`B01001_005`,
                         A.`B01001_006`,
                         A.`B01001_007`,
                         A.`B01001_008`,
                         A.`B01001_009`,
                         A.`B01001_010`,
                         A.`B01001_011`,
                         A.`B01001_012`,
                         A.`B01001_013`,
                         A.`B01001_014`,
                         A.`B01001_015`,
                         A.`B01001_016`,
                         A.`B01001_017`,
                         A.`B01001_018`,
                         A.`B01001_019`,
                         A.`B01001_020`,
                         A.`B01001_021`,
                         A.`B01001_022`,
                         A.`B01001_023`,
                         A.`B01001_024`,
                         A.`B01001_025`,
                         A.`B01001_026`,
                         A.`B01001_027`,
                         A.`B01001_028`,
                         A.`B01001_029`,
                         A.`B01001_030`,
                         A.`B01001_031`,
                         A.`B01001_032`,
                         A.`B01001_033`,
                         A.`B01001_034`,
                         A.`B01001_035`,
                         A.`B01001_036`,
                         A.`B01001_037`,
                         A.`B01001_038`,
                         A.`B01001_039`,
                         A.`B01001_040`,
                         A.`B01001_041`,
                         A.`B01001_042`,
                         A.`B01001_043`,
                         A.`B01001_044`,
                         A.`B01001_045`,
                         A.`B01001_046`,
                         A.`B01001_047`,
                         A.`B01001_048`,
                         A.`B01001_049`,
                         B.`B19062_001`,
                         B.`B19001_002`, 
                         B.`B19001_003`,
                         B.`B19001_004`, 
                         B.`B19001_005`, 
                         B.`B19001_006`, 
                         B.`B19001_007`, 
                         B.`B19001_008`, 
                         B.`B19001_009`, 
                         B.`B19001_010`, 
                         B.`B19001_011`, 
                         B.`B19001_012`, 
                         B.`B19001_013`, 
                         B.`B19001_014`, 
                         B.`B19001_015`, 
                         B.`B19001_016`,
                         B.`B19001_017`
                         from `uscensusbureau`.`acs-2015-5-e-agesex`.`USA_All_States` as A,
                         `uscensusbureau`.`acs-2015-5-e-income`.`USA_All_States` as B
                         where A.AreaName = B.AreaName"
                         )
```

We then took the data table related to Facebook traffic penetration.

```{r, warning=FALSE}

df1Second <- data.world::query(connection = conn,
                               type = "sql",
                               dataset = "thule179/s-17-dv-final-project",         
                               "select
                               C.`Population (2010 Est.)`,
                               C.`Population % of USA`,
                               C.`Internet users June, 2010`,
                               C.`Internet Penetration`,
                               C.`Facebook users August, 2010`,
                               C.`Facebook Penetration`
                               from Facebook as C")
```

Lastly, we brought in data points for internet usage and connectivity.

```{r, warning=FALSE}
df_InternetConnectivity <- data.world::query(connection = conn, type = "sql",
                                             dataset = "thule179/s-17-dv-final-project",
                                             "SELECT c.State as State, `Internet Connectivity`.`No connection anywhere (%)` as NoConnectionAnywhere, `Internet Connectivity`.`No home connection, but connect elsewhere (%)`
as NoHomeConnection_ConnectElseWhere, `Internet Connectivity`.`Connect at home only (%)` as ConnectAtHomeOnly, 
                                          InternetUsageAtHome.`Internet Usage At Home` as InternetUsageAtHome, InternetUsageAtWork.`Internet Usage At Work` as InternetUsageAtWork,
                                          InternetUsageAtCoffeeShops.InternetUsageAtCoffeeShops as InternetUsageAtCoffeeShops,
                                          InternetUsage.InternetUsage as InternetUsage
                                          FROM `Internet Connectivity.xlsx/Internet Connectivity`as c, `Internet Connectivity.xlsx/InternetUsageAtHome` as h,
                                          InternetUsageAtCoffeeShops as s, InternetUsageAtWork as w, InternetUsage as i
                                          where c.State = h.State and s.State = h.State and h.State = w.State and w.State = i.State")
```

## 2. Mapping the data

To reduce the number of columns in our main table and to make the data easier for analysis, we categorized the US population into age and income groups, separated by gender.

### a. Grouping By Age
Put males into age categories.
```{r,warning=FALSE}
male0to9 <- df1$B01001_003 + df1$B01001_004
male10to19 <- df1$B01001_005 + df1$B01001_006 + df1$B01001_007
male20to29 <- df1$B01001_008 + df1$B01001_009 + df1$B01001_010 + df1$B01001_011
male30to39 <- df1$B01001_012 + df1$B01001_013
male40to49 <- df1$B01001_014 + df1$B01001_015
male50to59 <- df1$B01001_016 + df1$B01001_017
male60to69 <- df1$B01001_018 + df1$B01001_019 + df1$B01001_020 + df1$B01001_021
male70to79 <- df1$B01001_022 + df1$B01001_023
male80andUp <- df1$B01001_024 + df1$B01001_025
```

Put females into age categories.
```{r,warning=FALSE}
female0to9 <- df1$B01001_027 + df1$B01001_028
female10to19 <- df1$B01001_029 + df1$B01001_030 + df1$B01001_031
female20to29 <- df1$B01001_032 + df1$B01001_033 + df1$B01001_034 + df1$B01001_035
female30to39 <- df1$B01001_036 + df1$B01001_037
female40to49 <- df1$B01001_038 + df1$B01001_039
female50to59 <- df1$B01001_040 + df1$B01001_041
female60to69 <- df1$B01001_042 + df1$B01001_043 + df1$B01001_044 + df1$B01001_045
female70to79 <- df1$B01001_046 + df1$B01001_047
female80andUp <- df1$B01001_048 + df1$B01001_049
```

### b. Grouping By Income
Group income into categories.
```{r,warning=FALSE}
Agg_Income <- df1$B19062_001
TenToThirtyK <- df1$B19001_002 + df1$B19001_003 + df1$B19001_004 + df1$B19001_005 + df1$B19001_006
ThirtyToFiftyK <- df1$B19001_007 + df1$B19001_008 + df1$B19001_009 + df1$B19001_010
FiftyToHundredK <- df1$B19001_011 + df1$B19001_012 + df1$B19001_013
HundredToHundredFiftyK <- df1$B19001_014 + df1$B19001_015
HundredFiftyPlus <- df1$B19001_016 + df1$B19001_017
```

### c. Binding the data
Bind all the categories into a data frame.
```{r,warning=FALSE}
df2 <- as.data.frame(cbind(State = df1$AreaName, TotalPopulation = df1$B01001_001, male0to9, male10to19, male20to29, male30to39, male40to49, male50to59, male60to69, male70to79, male80andUp, female0to9, female10to19, female20to29, female30to39, female40to49, female50to59, female60to69, female70to79, female80andUp, Agg_Income,TenToThirtyK, ThirtyToFiftyK, FiftyToHundredK, HundredToHundredFiftyK, HundredFiftyPlus), stringsAsFactors = FALSE)
```

Change the numeric columns to numeric.
```{r,warning=FALSE}
df2[-1] <- as.data.frame(apply(df2[-1], 2, as.numeric))
df2 <- cbind(df2, df1Second)
```

Do an inner join by State of df2 and InternetConnectivity.
```{r,warning=FALSE}
df2 <- merge(df2,df_InternetConnectivity, by ="State")
```

## 3. Cleaning the Data
Now we have our main data table, we'd need to clean up our data to prepare for analysis.

Get average income for each state.
```{r,warning=FALSE}
df2$PerCapitaIncome <-  df2$Agg_Income / df2$TotalPopulation
```

Remove '\n' line breaks and commas in rows to convert InternetUsage into numbers.
```{r,warning=FALSE}
df2[,'InternetUsage'] <- gsub(",","",df2[,'InternetUsage'])
df2[,'InternetUsage'] <- gsub("\n","",df2[,'InternetUsage'])
df2[,'InternetUsageAtCoffeeShops'] <- gsub(",","",df2[,'InternetUsageAtCoffeeShops'])
df2[,'InternetUsageAtCoffeeShops'] <- gsub("\n","",df2[,'InternetUsageAtCoffeeShops'])
df2[,'InternetUsageAtWork'] <- gsub(",","",df2[,'InternetUsageAtWork'])
df2[,'InternetUsageAtWork'] <- gsub("\n","",df2[,'InternetUsageAtWork'])
df2[,'InternetUsageAtHome'] <- gsub(",","",df2[,'InternetUsageAtHome'])
df2[,'InternetUsageAtHome'] <- gsub("\n","",df2[,'InternetUsageAtHome'])
df2$InternetUsage <- as.numeric(df2$InternetUsage)
df2$InternetUsageAtCoffeeShops <-  as.numeric(df2$InternetUsageAtCoffeeShops)
df2$InternetUsageAtWork <-  as.numeric(df2$InternetUsageAtWork)
df2$InternetUsageAtHome <-  as.numeric(df2$InternetUsageAtHome)
```

Get average internet usage per person for each State.
```{r,warning=FALSE}
df2$Avg_InternetUsage <- df2$InternetUsage
df2$Avg_InternetUsage <- df2$InternetUsage / df2$TotalPopulation
```

Get average internet usage at work.
```{r,warning=FALSE}
df2$Avg_InternetUsageAtWork <- df2$InternetUsageAtWork
df2$Avg_InternetUsageAtWork <- df2$InternetUsageAtWork / df2$TotalPopulation
```

Get average internet usage at home.
```{r,warning=FALSE}
df2$Avg_InternetUsageAtHome <- df2$InternetUsageAtHome
df2$Avg_InternetUsageAtHome <- df2$InternetUsageAtHome / df2$TotalPopulation
```

Get average internet usage at coffee shops.
```{r,warning=FALSE}
df2$Avg_InternetUsageAtCoffeeShops <- df2$InternetUsageAtCoffeeShops
df2$Avg_InternetUsageAtCoffeeShops <- df2$InternetUsageAtCoffeeShops / df2$TotalPopulation
```

Find high, medium, and low ranges of Internet Usage.
```{r,warning=FALSE}
sorted_df2 <- df2[order(df2$Avg_InternetUsage),] # sort df2 from lowest to highest based on avg_internet usage
low_range <- c(sorted_df2$Avg_InternetUsage[c(0: (51/3))]) # subset states with low Internet Usage
medium_range <- c(sorted_df2$Avg_InternetUsage[c((51/3): (2*51/3))]) # subset states with medium Internet Usage
high_range <- c(sorted_df2$Avg_InternetUsage[c((2*51/3) : 51)]) # subset states with high Internet Usage

# Find high, medium, and low ranges of Internert usage at work
sorted_df2_work <- df2[order(df2$Avg_InternetUsageAtWork ),]
low_range_work <- c(sorted_df2_work$Avg_InternetUsageAtWork[c(0: (51/3))]) 
medium_range_work <- c(sorted_df2_work$Avg_InternetUsageAtWork[c((51/3): (2*51/3))])
high_range_work <- c(sorted_df2_work$Avg_InternetUsageAtWork[c((2*51/3) : 51)])

# Find high, medium, and low ranges of Internert usage at home
sorted_df2_home <- df2[order(df2$Avg_InternetUsageAtHome ),] 
low_range_home <- c(sorted_df2_home$Avg_InternetUsageAtHome[c(0: (51/3))]) 
medium_range_home <- c(sorted_df2_home$Avg_InternetUsageAtHome[c((51/3): (2*51/3))])
high_range_home <- c(sorted_df2_home$Avg_InternetUsageAtHome[c((2*51/3) : 51)]) 

# Find high, medium, and low ranges of Internert usage at coffee shops
sorted_df2_coffee <- df2[order(df2$Avg_InternetUsageAtCoffeeShops),] 
low_range_coffee <- c(sorted_df2_coffee$Avg_InternetUsageAtCoffeeShops[c(0: (51/3))]) 
medium_range_coffee <- c(sorted_df2_coffee$Avg_InternetUsageAtCoffeeShops[c((51/3): (2*51/3))])
high_range_coffee <- c(sorted_df2_coffee$Avg_InternetUsageAtCoffeeShops[c((2*51/3) : 51)]) 

# Create column for Internet Usage Levels
df2$InternetUsageLevel <- df2$InternetUsage
df2$InternetUsageAtHomeLevel <- df2$InternetUsageAtHome
df2$InternetUsageAtWorkLevel <- df2$InternetUsageAtWork
df2$InternetUsageAtCoffeeShopsLevel<- df2$InternetUsageAtCoffeeShops
```

Create a function to categorize internet usage level.
```{r,warning=FALSE}
UsageLevel <- function(col, low_range, medium_range, high_range ){
  for (i in 1:nrow(df2)){
    if(col[i] %in% low_range){
      col[i] = "Low"
    }
    if(col[i] %in% medium_range){
      col[i] = "Medium"
    }
    if(col[i] %in% high_range){
      col[i] = "High"
    }
  }

  result <- col
  return(result)
}
```

Get internet usage level by category.
```{r,warning=FALSE}
df2$InternetUsageLevel <- UsageLevel(df2$Avg_InternetUsage,low_range,medium_range,high_range)
df2$InternetUsageAtHomeLevel <- UsageLevel(df2$Avg_InternetUsageAtHome, low_range_home, medium_range_home, high_range_home)
df2$InternetUsageAtWorkLevel <- UsageLevel(df2$Avg_InternetUsageAtWork, low_range_work, medium_range_work, high_range_work)
df2$InternetUsageAtCoffeeShopsLevel <- UsageLevel(df2$Avg_InternetUsageAtCoffeeShops, low_range_coffee, medium_range_coffee, high_range_coffee)
```

Add percent of total population for each age range.
```{r,warning=FALSE}
df2$YoungProportion <- (df2$male0to9 + df2$male10to19 + df2$male20to29 + df2$female0to9 + df2$female10to19 + df2$female20to29) / df2$TotalPopulation
df2$MiddleProportion <- (df2$male30to39 + df2$male40to49 + df2$male50to59 + df2$female30to39 + df2$female40to49 + df2$female50to59) / df2$TotalPopulation
df2$OldProportion <- 1 - df2$YoungProportion - df2$MiddleProportion
```

Create discrete categories for Young Proportion.
```{r,warning=FALSE}
third = quantile(df2$YoungProportion, 1/3)
two_third = quantile(df2$YoungProportion, 2/3)
df2$YoungCategories <- if_else(df2$YoungProportion < third, "Low", if_else(df2$YoungProportion < two_third, "Medium", "High"))
```

Export to csv.
```{r,warning=FALSE}
write.csv(df2,file="./CleanedInternetUsageByState.csv")
```

Now we have a nicely formatted csv file ready for analysis!

# IV. Data Analysis
## Key Findings:

### Low internet usage in Southern area of United States
+ The first interesting analysis we discovered was a trend of **low internet usage at both home and work** in the southern and southwestern regions of the United States. Similarly, there is **high internet usage levels at home and work** in the northeastern area of the United States as well as the area from Minnesota to Illinois. In general, it seems like the **further north you are in the United States, the higher internet usage levels there are.** This may be due to the fact that people in the South are more likely to live and work in rural lifestyles than the northern states. Note that darker colors represent higher internet usage rates.
```{r, message=FALSE, echo = FALSE, fig.width=10, fig.height = 10, fig.align='center'}
df3 <- data.frame(state = df2$State, interaction(df2$InternetUsageAtHomeLevel, df2$InternetUsageAtWorkLevel))
names(df3) <- c("region", "value")
df3$value <- factor(df3$value, levels = c("Low.Low", "Low.Medium", "Low.High", "Medium.Low", "Medium.Medium", "Medium.High", "High.Low", "High.Medium", "High.High"))
df3$region <- tolower(df3$region)
levels(df3$value) <- c("Low Home Usage/Low Work Usage", "Low Home Usage/Medium Work Usage", "Low Home Usage/High Work Usage", "Medium Home Usage/Low Work Usage", "Medium Home Usage/Medium Work Usage", "Medium Home Usage/High Work Usage", "High Home Usage/Low Work Usage", "High Home Usage/Medium Work Usage", "High Home Usage/High Work Usage")
data("continental_us_states")
c1 <- state_choropleth(df3,
                      zoom = continental_us_states)
c1 + scale_fill_brewer(type = "seq", palette = 12) + labs(fill = "Internet Usage (Home/Work)")

```

*****
### Internet Usage vs Young People Population
+ After viewing this choroplethr map and seeing areas with high and low internet usage, we decided to make a bar chart to view overall internet trends, including the states with the high overall internet usage. As seen in the previous map, **Vermont, New Hampshire, Minnesota, and Wisconsin have the highest overall internet usage, while Mississippi, Alabama, Tennessee, and Hawaii have the lowest overall internet usage.**

```{r, message=FALSE, echo = FALSE, fig.width=10, fig.height = 10, fig.align='center'}
df3 <- df2%>%select(State,Avg_InternetUsage)
df3$InternetUsage_z <-  round((df3$Avg_InternetUsage - mean(df3$Avg_InternetUsage))/sd(df3$Avg_InternetUsage), 2)  # compute normalized Internet Usage
df3$InternetUsage_type <- ifelse(df3$InternetUsage_z < 0, "below", "above")  # above / below avg flag
df3 <- df3[order(df3$InternetUsage_z), ]  # sort
df3$State <- factor(df3$State, levels = unique(df3$State)[order(df3$InternetUsage_z)]) 
                  
InternetUsage_bar <- ggplot(df3, aes(x=State, y=InternetUsage_z, label=InternetUsage_z)) + geom_bar(stat='identity', aes(fill=InternetUsage_type), width=.5)  +scale_fill_manual(name="Internet Usage", labels = c("Above Average", "Below Average"), values = c("above"="#00ba38", "below"="#f8766d")) + coord_flip()
InternetUsage_bar
```
In order to further analyse these results, we can use the census data to view the proportion of young people for the Top 3 and Bottom 3 states by internet usage. Note that **the top two states by internet usage (New Hampshire, Vermont) have much lower proportions of young people than the bottom states (Alabama, Mississippi, Tennessee)**.

```{r, message=FALSE, echo = FALSE, fig.width=10, fig.height = 10, fig.align='center'}
df4 <- df2%>%select(State,YoungProportion)
df4$YoungPeople_z <-  round((df4$YoungProportion - mean(df4$YoungProportion))/sd(df4$YoungProportion), 2)  # compute normalized Young Proportion
df4$YoungPeople_type <- ifelse(df4$YoungPeople_z < 0, "below", "above")  # above / below avg flag
df4 <- df4[order(df4$YoungPeople_z), ]  # sort
df4$State <- factor(df4$State, levels = df4$State) 
df5 <- df4[df4$State %in% c("Vermont", "New Hampshire", "Minnesota", "Mississippi", "Alabama", "Tennessee"),]

YoungPeople_bar <- ggplot(df5, aes(x=State, y=YoungPeople_z, label=YoungPeople_z)) + geom_bar(stat='identity', aes(fill=YoungPeople_type), width=.5)  +scale_fill_manual(name="Young Proportion", labels = c("Above Average", "Below Average"), values = c("above"="#00ba38", "below"="#f8766d")) + coord_flip()
YoungPeople_bar

```

***
### 3-D Scatterplot - Young People vs Income vs Internet Usage
+ Finally, in order to get a look at the data from all angles, we created a 3D Scatterplot in plotly with the proportion of Young People vs Median Income vs Internet Usage. Notice there is an extreme outlier (drag and drop the graph to see it better). This chart shows **how District of Colombia is drastically different from the 50 states in terms of young adult percentage.** It also shows that 
**median income is positively correlated with internet usage.**

```{r, message=FALSE, echo = FALSE, fig.width=4, fig.height = 4, fig.align='center'}

df2$adult20to29 <- df2$male20to29 + df2$female20to29
df2$youngAdultPercentage <- df2$adult20to29 / df2$TotalPopulation
normalize <- function(x){
  return((x - mean(x)) / sd(x))
}
df2$normalizedIncome <- normalize(df2$PerCapitaIncome)
p <- plot_ly(x = ~list(df2$Avg_InternetUsage), 
             y = ~list(df2$normalizedIncome), 
             z = ~list(df2$youngAdultPercentage),
             marker = list(color = ~mpg, colorscale = c('#FFE1A1', '#683531'), showscale = FALSE), 
             hoverinfo = 'text', 
             text = ~paste('Internet Usage (Avg) : ', df2$Avg_InternetUsage, 
                           '</br>Per Capita Income : ', df2$PerCapitaIncome,
                           '</br> Young Adult Percentage : ', df2$youngAdultPercentage,
                           '</br> State : ', df2$State), color = df2$State) %>% 
             add_markers() %>% 
             layout(scene = list(xaxis = list(title = 'Average Internet Usage'),
                                      yaxis = list(title = 'Median Income'),
                                      zaxis = list(title = 'Young Adult Percentage')))
p
```
